Coalescent inference of mutation spectrum histories from sample frequency spectra

### 2019-08-16 manuscript/main.tex
much math tidying

Suppose $n$ haplotypes are sampled in the present, and let rvs $T = T_2,\dots,T_n$ denote the coalescent times measured retrospectively from the present (i.e. $T_n$ is the most recent coalescent time, and $T_2$ is the TMRCA of the sample).
As in \cite{Griffiths1998-qf} \S3, we consider a marked Poisson process in which every mutation is assigned a random label drawn iid from the uniform distribution on $(0,1)$.
This is tantamount to an (uncountably) infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
Further suppose that mutation rate is not constant, but a specified function of time $\mu(t) > 0$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present).
A given line in the coalescent tree then acquires mutations on a genomic region $M\subset(0,1)$ at rate $|M|\mu(t)$, where $|M|$ is the Lebesgue measure of $M$ (e.g.\ if $M$ is a subinterval, $|M|$ is the length of $M$).

Let $C_h$ denote the event that there is a mutation with label $U\in(x,x+h)\subseteq(0,1)$ that subtends $b$ haplotypes in the sample, and let $I_k$ denote the event that this mutation arose at a time when the sample had $k$ ancestral lines.
For small $h$ we then have
\begin{align*}
\mathbbm{P}(C_h\mid T) &= \sum_{k=2}^n \mathbbm{P}(C_h, I_k\mid T) = \sum_{k=2}^n p_{n,k}(b)\mathbbm{P}(I_k, U\in(x,x+h)\mid T)\\
&= \sum_{k=2}^n p_{n,k}(b)\left(kh\int_{T_{k-1}}^{T_k}\mu(t)dt + O(h)\right),
\end{align*}
where $p_{n,k}(b) = \binom{n-b-1}{k-2}/\binom{n-1}{k-1}$ is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9).
Marginalizing $T$ gives
\begin{align*}
\mathbbm{P}(C_h) &\simeq h\sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right] = h\sum_{k=2}^n p_{n,k}(b)\ k\ \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right),
\end{align*}
where $\nu(t)\equiv\int_0^t\mu(t)dt$.
% Now, since $\mu(t) > 0$ we have $\nu(t) \nearrow$.
% With $\pi_k(t_k)$ denoting the marginal density of $T_k$, define rv $X_k \equiv \nu(T_k)$, which has density $f_{X_k}(x_k) = \pi_k(\nu^{-1}(x_k))$.
Using eqns 5-8 from \cite{Polanski2003-kg} we can write
\begin{align*}
\mathbbm{E}\left[\nu(T_{k})\right] &= \sum_{j=k}^n A_j^k\int_0^\infty\nu(t)q_j(t)dt
\end{align*}
where
\begin{align*}
A_j^k &\equiv \frac{\prod_{l=k\ne j}^{n}\binom{l}{2}}{\prod_{l=k\ne j}^{n}\left[\binom{l}{2}-\binom{j}{2}\right]}, k\le j\le n,\\
A_n^n &\equiv 1,\\
q_j(t) &\equiv \frac{\binom{j}{2}}{\eta(t)}\exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right],
\end{align*}
and $\eta(t)$ is the haploid effective population size history.
Note that $q_j(t)$ is the density of the time to the first coalescent event among any subset of $j$ individuals in the present.
This can be made explicit by defining the inhomogeneous Poisson process intensity function $\lambda_j(t) = \binom{j}{2}/\eta(t)$.

The $b$-th component of the expected SFS is then

\begin{align*}
\xi_b &= \sum_{k=2}^n p_{n,k}(b) k \left(\mathbbm{E}\left[\nu(T_{k})\right] - \mathbbm{E}\left[\nu(T_{k-1})\right]\right)\\
&= \sum_{k=2}^n p_{n,k}(b) k \sum_{j=k-1}^{n}B_j^k\int_0^\infty\nu(t)q_j(t)dt
\end{align*}
where
\[
B_j^k \equiv
\begin{cases}
A_{k-1}^{k-1},\ j=k-1\\
A_j^k - A_j^{k-1},\ k\le j\le n\\
\end{cases}
\]
Using the definition of $\nu$ and exchanging integration order gives
\begin{align*}
\int_0^\infty\nu(t)q_j(t)dt &= \int_0^\infty \left(\int_0^t\mu(x)dx\right) q_j(t)dt\\
&= \int_0^\infty \int_0^\infty 1_{[0<x<t]}(x)\mu(x)dx\ q_j(t)dt\\
&= \int_0^\infty \left(\int_0^\infty 1_{[0<x<t]}(x)q_j(t)dt\right)\mu(x) dx\\
&= \int_0^\infty \left(1-\int_0^x q_j(t)dt\right)\mu(x) dx\\
&= \int_0^\infty \left(1-Q_j(t)\right)\mu(x) dx,
\end{align*}
where $Q_j$ is the distribution function corresponding to density $q_j$.

We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
\[
\xi_b = \left<\psi_b, \mu\right>,
\]
where
\begin{align*}
\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k \left(1-Q_j(t)\right).
\end{align*}

>>>>

Suppose $n$ haplotypes are sampled in the present, and let rvs $T = T_2,\dots,T_n$ denote the coalescent times measured retrospectively from the present (i.e. $T_n$ is the most recent coalescent time, and $T_2$ is the TMRCA of the sample).
As in \cite{Griffiths1998-qf} \S3, we consider a marked Poisson process in which every mutation is assigned a random label drawn iid from the uniform distribution on $(0,1)$.
This is tantamount to an (uncountably) infinite sites assumption, with the unit interval representing the genome, and the random variate labels representing mutant sites.
Further suppose that mutation rate is not constant, but a specified function of time $\mu(t) \ge 0$ (measured in mutations per genome per generation) applying equally to all lines in the coalescent tree at a given time $t$ (measured retrospectively from the present in units of Wright-Fisher generations).
A given line in the coalescent tree then acquires mutations on a genomic subinterval $(x,x+\delta x)$ at rate $\mu(t)\delta x$.
The following argument can be generalized to a nonuniform distribution of mutation intensity over the genome without changing the results.

Let rv $M_{\delta x, b}$ denote the event that a mutation subtending $b$ haplotypes in the sample occurred within a given genomic interval $(x,x+\delta x)$ (identical for all $x$, given the uniform labeling distribution).
Let rv $M_{\delta x, b, k}$ denote the event that this mutation occurred at a time when the sample had $k$ ancestral lines (during the $k$th intercoalescent time interval).
For small $\delta x$ and finite $\mu(t)$ we have
\begin{align*}
\mathbbm{P}(M_{\delta x, b}\mid T) &= \sum_{k=2}^n \mathbbm{P}(M_{\delta x, b, k}\mid T)\\
&= \sum_{k=2}^n p_{n,k}(b)\left(k\delta x\int_{T_{k-1}}^{T_k}\mu(t)dt + O(\delta x)\right),
\end{align*}
where
\begin{equation}
\label{eqn:p}
p_{n,k}(b) = \frac{\binom{n-b-1}{k-2}}{\binom{n-1}{k-1}}
\end{equation}
is the probability that a mutant that arose when there were $k$ ancestral lines of $n$ sampled haplotypes will be present in $b$ of them (see \cite{Griffiths1998-qf} eqn 1.9), and the quantity in parenthesis is the probability that a mutation arose during the $k$th intercoalescent interval in a small genomic interval of size $\delta x$.
Marginalizing $T$ gives
\begin{align*}
\mathbbm{P}(M_{\delta x, b}) &\simeq \delta x\sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right].
\end{align*}
For small $\delta x$, each such interval contains zero or one mutations, so the expected number of mutations subtending $b$ haplotypes (i.e.\ the $b$th component of the SFS) is
\begin{align}
\label{eqn:xi}
\xi_b &= \int d\mathbbm{P}(M_{dx, b}) = \sum_{k=2}^n p_{n,k}(b)\ k\ \mathbbm{E}_T\left[\int_{T_{k-1}}^{T_k}\mu(t)dt\right]\nonumber\\
&= \sum_{k=2}^n p_{n,k}(b)\ k\ \left(\mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] - \mathbbm{E}_{T_{k-1}}\left[\int_0^{T_{k-1}}\mu(t)dt\right]\right)
\end{align}

\cite{Polanski2003-kg} (eqns 5-8) give the marginal density for the coalescent time $T_k$ as
\begin{equation}
\label{eqn:pi}
\pi_k(t_k) = \sum_{j=k}^n A_j^k q_j(t_k)
\end{equation}
where
\begin{align*}
A_j^k &\equiv \frac{\prod_{l=k\ne j}^{n}\binom{l}{2}}{\prod_{l=k\ne j}^{n}\left[\binom{l}{2}-\binom{j}{2}\right]}, k\le j\le n,\\
A_n^n &\equiv 1,\\
q_j(t) &\equiv \frac{\binom{j}{2}}{\eta(t)}\exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right],
\end{align*}
and $\eta(t)$ is the haploid effective population size history.
Note that $q_j(t)$ is the density of the time to the first coalescent event among any subset of $j$ individuals in the present, with inhomogeneous Poisson intensity function $\binom{j}{2}/\eta(t)$.

The expectations in \eqref{eqn:xi} can be expressed using \eqref{eqn:pi} as
\begin{align}
\label{eqn:exp}
\mathbbm{E}_{T_k}\left[\int_0^{T_k}\mu(t)dt\right] &= \int_0^\infty\pi_k(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
&= \sum_{j=k}^n A_j^k\int_0^\infty q_j(t_k)\int_0^{t_k}\mu(t)dt dt_k\nonumber\\
&= \sum_{j=k}^n A_j^k\int_0^\infty r_j(t)\mu(t)dt
\end{align}
where in the last line we've exchanged integration order and defined the inhomogeneous Poisson survival function
\begin{equation}
\label{eqn:r}
r_j(t) \equiv \exp\left[-\binom{j}{2}\int_0^t\frac{dx}{\eta(x)}\right]
\end{equation}
corresponding to density $q_j(t)$.
Using \eqref{eqn:exp} in \eqref{eqn:xi} gives
\begin{align}
\label{eqn:xi2}
\xi_b &= \sum_{k=2}^n p_{n,k}(b)\ k\ \left(\sum_{j=k}^n A_j^k\int_0^\infty r_j(t)\mu(t)dt - \sum_{j=k-1}^n A_j^{k-1}\int_0^\infty r_j(t)\mu(t)dt\right)\nonumber\\
&= \sum_{k=2}^n p_{n,k}(b) k \sum_{j=k-1}^n B_j^k\int_0^\infty r_j(t)\mu(t)dt
\end{align}
where
\[
B_j^k \equiv
\begin{cases}
A_{k-1}^{k-1},\ j=k-1\\
A_j^k - A_j^{k-1},\ k\le j\le n\\
\end{cases}
\]

We then have a manifestly linear expression for the expected SFS as a function of the mutation rate history $\mu(t)$:
\[
\xi_b = \left<\psi_b, \mu\right>,
\]
where
\begin{align*}
\psi_b(t) &\equiv \sum_{k=2}^n p_{n,k}(b)\ k \sum_{j=k-1}^{n}B_j^k r_j(t).
\end{align*}
